Notes on generating initial models for mapping into {\tt MAESTRO}

### 2008-08-27 docs/initial_models/paper.tex
some more work... my multilevel notes aren't quite right in section 3 yet.  some sublte tweaking may have to be done in the algorithm as well

\section{Creating the Model Data from Raw Data}

@@

We need to generate model data with the same resolution as the finest base state resolution.

@@

In {\tt base\_state.f90}, using the subroutine {\tt init\_base\_state.f90} (which is actually a terrible name since we are setting the initial data, not the base state), we set the initial data $\rho^{\init}, T^{\init}, p^{\init}$ and $X^{\init}$ equal to the model data.  Then, we set $h^{\init} = h^{\init}(\rho^{\init},T^{\init},X^{\init})$.  Note that in the code, $p^{\init}$ is also overwritten, but the value of $p^{\init}$ does not change since we called $p^{\model} = p^{\model}(\rho^{\model},T^{\model},X^{\model})$ at the end of the previous fortran subroutine.

@@

For the finest level, we can set the initial data equal to the model data.  For the non-finest levels, we use linear interpolation to compute the initial data.  For cells that are covered by finer cells, we average $\rho^{\init}, (\rho h)^{\init},$ and $(\rho X)^{\init}$ from the fine cell to get the coarse cell value.  Then, for every cell not at the finest level, we call the EOS to compute $p^{\init},T^{\init} = p^{\init},T^{\init}(\rho^{\init},h^{\init},X^{\init})$.\MarginPar{Now $p^{\init}$ and $\rho^{\init}$ are not in HSE.  Is this a problem?}

@@

\item For plane-parallel problems, fill $\rho^{\init}, h^{\init}, X^{\init}$, and $T^{\init}$ onto a multifab to obtain the full state $\rho, h, X$, and $T$.  For spherical problems, fill $\rho^{\init}, p^{\init}$, and $X^{\init}$ onto a multifab to obtain the full state $\rho, p$, and $X$, and then call the EOS to obtain the full state $h,T = h,T(\rho,p,X)$.  Note that the multifab version of $p$ is temporary, and can be deallocated after this step.

@@

Refer to figure \ref{Fig:Multilevel Initialization}.  The dots represent the initial data, which was generated at the finest resolution on the domain, and is in HSE as well as thermodynamic equilibrium.  The circles represent the fine level data points in the 1D array.  The squares represent the coarse level data points in the 1D array.  We modify step 1 in \S~\ref{Sec:Creating the Base State and Full State} as follows:
\item Step 1
\item Copy the initial data from the dots directly into the circles.  For the squares, compute only $\rho$ and $X$ by using linear interpolation from the two nearest corresponding dots.
\item Integrate to get $p$ in the squares.  We need the fancy stencils at coarse-fine interfaces in equations (\ref{Coarse-Fine Stencil}) and (\ref{Fine-Coarse Stencil}).  Away from coarse-fine interfaces, we can use our standard HSE discretization.
\item Call EOS to get $T$ and $h$ in the squares, i.e., $T,h = T,h(\rho,p,X)$.
\item We now copy this 1D profile onto the full multifab.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tpb]
\begin{center}
\epsffile{./multilevel_init.eps}
\caption{Multilevel Initialization}
\label{Fig:Multilevel Initialization}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Coarse-Fine HSE Discretization}
Integrating the HSE discretization upward, when we encounter a change from coarse to fine, here is an equation to integrate up to the coarse-fine interface ($l$ is the ``finer'' level, $l-1$ is the ``coarser'' level)
\frac{p_{r-\myhalf}^l - p_{\sfrac{r}{2}-1}^{l-1}}{\Delta r^{l-1}/2} = \frac{\rho_{r-\myhalf}^l + \rho_{\sfrac{r}{2}-1}^{l-1}}{2}g ~~~ \rightarrow ~~~ p_{r-\myhalf}^l = p_{\sfrac{r}{2}-1}^{l-1} + \frac{\Delta r^{l-1} g}{4}\left(\rho_{r-\myhalf}^l + \rho_{\sfrac{r}{2}-1}^{l-1}\right).
Here is an equation to integrate from the coarse-fine interface up to the cell center
\frac{p_r^l - p_{r-\myhalf}^l}{\Delta r^l/2} = \frac{\rho_r^l + \rho_{r-\myhalf}^l}{2}g ~~~ \rightarrow ~~~ p_r^l = p_{r-\myhalf}^l + \frac{\Delta r^l g}{4}\left(\rho_r^l + \rho_{r-\myhalf}^l\right).
Combining equations gives
p_r^l = p_{\sfrac{r}{2}-1}^{l-1} + \frac{\Delta r^{l-1} g}{4}\left(\rho_{r-\myhalf}^l + \rho_{\sfrac{r}{2}-1}^{l-1}\right) + \frac{\Delta r^l g}{4}\left(\rho_r^l + \rho_{r-\myhalf}^l\right).
We can simplify using
\Delta r^{l-1} = 2\Delta r^l,
\rho_{r-\myhalf}^l = \frac{2}{3}\rho_r^l + \frac{1}{3}\rho_{\sfrac{r}{2}-1}^{l-1}.
Simplifying
p_r^l &=& p_{\sfrac{r}{2}-1}^{l-1} + \frac{\Delta r^l g}{2}\left(\frac{2}{3}\rho_r^l + \frac{1}{3}\rho_{\sfrac{r}{2}-1}^{l-1} + \rho_{\sfrac{r}{2}-1}^{l-1}\right) + \frac{\Delta r^l g}{4}\left(\rho_r^l + \frac{2}{3}\rho_r^l + \frac{1}{3}\rho_{\sfrac{r}{2}-1}^{l-1}\right) \nonumber \\
&=& p_{\sfrac{r}{2}-1}^{l-1} + \frac{3\Delta r^l g}{4}\left(\rho_{\sfrac{r}{2}-1}^{l-1} + \rho_r^l\right).\label{Coarse-Fine Stencil}
When we encounter a change from fine to coarse, analogously
p_{(r+1)/2}^{l-1} = p_{r}^l + \frac{3\Delta r^l g}{4}\left(\rho_{r}^l+\rho_{(r+1)/2}^{l-1}\right).\label{Fine-Coarse Stencil}
\end{document}

>>>>

\section{Creating the Model Data from Raw Data}\label{Sec:Creating the Model Data from Raw Data}

@@

We do the same thing as the single-level case except that we generate model data with the same resolution as the finest base state resolution.

@@

In {\tt base\_state.f90}, using the subroutine {\tt init\_base\_state.f90} (which is actually a terrible name since we are setting the initial data, not the base state), we set the initial data $\rho^{\init}, T^{\init}, p^{\init}$ and $X^{\init}$ equal to the model data.  Then, we set $h^{\init} = h^{\init}(\rho^{\init},T^{\init},X^{\init})$.  Note that in the code, $p^{\init}$ is also overwritten, but the value of $p^{\init}$ does not change since we called $p^{\model} = p^{\model}(\rho^{\model},T^{\model},X^{\model})$ at the end of the fortran subroutine in \S~\ref{Sec:Creating the Model Data from Raw Data}.

@@

We do the same thing as the single-level case except that we generage initial data at the non-finest levels by using linear interpolation from the two nearest model points.  Then, after we compute $h^{\init} = h^{\init}(\rho^{\init},T^{\init},X^{\init})$, for each cell that is covered by finer cells, we average $\rho^{\init}, (\rho h)^{\init},$ and $(\rho X)^{\init}$ from the fine cell to get the coarse cell value.  Then, for every cell not at the finest level, we call the EOS to compute $p^{\init},T^{\init} = p^{\init},T^{\init}(\rho^{\init},h^{\init},X^{\init})$.  Note that the initial data is not in HSE.  Later, when we compute the base state, we will make sure that the base state is in HSE.

@@

\item For plane-parallel problems, fill $\rho^{\init}, h^{\init}, X^{\init}$, and $T^{\init}$ onto a multifab to obtain the full state $\rho, h, X$, and $T$.  For spherical problems, fill $\rho^{\init}, p^{\init}$, and $X^{\init}$ onto a multifab to obtain the full state $\rho, p$, and $X$, and then call the EOS to obtain the full state $h,T = h,T(\rho,p,X)$.  Note that this particular multifab version of $p$ is temporary, and can be deallocated after this step.

@@

Refer to figure \ref{Fig:Multilevel Initialization}.  The dots represent the initial data, which was generated at the finest resolution on the domain, and is in HSE as well as thermodynamic equilibrium.  The circles represent the fine level data points in the 1D array.  The squares represent the coarse level data points in the 1D array.  We modify step 1 in \S~\ref{Sec:Creating the Base State and Full State} as follows:
\begin{enumerate}
\item Step 1
\begin{enumerate}
\item Copy the initial data from the dots directly into the circles.  For the squares, compute only $\rho$ and $X$ by using linear interpolation from the two nearest corresponding dots.
\item Integrate to get $p$ in the squares.  We need the fancy stencils at coarse-fine interfaces in equations (\ref{Coarse-Fine Stencil}) and (\ref{Fine-Coarse Stencil}).  Away from coarse-fine interfaces, we can use our standard HSE discretization.
\item Call EOS to get $T$ and $h$ in the squares, i.e., $T,h = T,h(\rho,p,X)$.
\item We now copy this 1D profile onto the full multifab.
\end{enumerate}
\end{enumerate}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[tpb]
\begin{center}
\epsffile{./multilevel_init.eps}
\caption{Multilevel Initialization}
\label{Fig:Multilevel Initialization}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsubsection{Coarse-Fine HSE Discretization}
Integrating the HSE discretization upward, when we encounter a change from coarse to fine, here is an equation to integrate up to the coarse-fine interface ($l$ is the ``finer'' level, $l-1$ is the ``coarser'' level)
\begin{equation}
\frac{p_{r-\myhalf}^l - p_{\sfrac{r}{2}-1}^{l-1}}{\Delta r^{l-1}/2} = \frac{\rho_{r-\myhalf}^l + \rho_{\sfrac{r}{2}-1}^{l-1}}{2}g ~~~ \rightarrow ~~~ p_{r-\myhalf}^l = p_{\sfrac{r}{2}-1}^{l-1} + \frac{\Delta r^{l-1} g}{4}\left(\rho_{r-\myhalf}^l + \rho_{\sfrac{r}{2}-1}^{l-1}\right).
\end{equation}
Here is an equation to integrate from the coarse-fine interface up to the cell center
\begin{equation}
\frac{p_r^l - p_{r-\myhalf}^l}{\Delta r^l/2} = \frac{\rho_r^l + \rho_{r-\myhalf}^l}{2}g ~~~ \rightarrow ~~~ p_r^l = p_{r-\myhalf}^l + \frac{\Delta r^l g}{4}\left(\rho_r^l + \rho_{r-\myhalf}^l\right).
\end{equation}
Combining equations gives
\begin{equation}
p_r^l = p_{\sfrac{r}{2}-1}^{l-1} + \frac{\Delta r^{l-1} g}{4}\left(\rho_{r-\myhalf}^l + \rho_{\sfrac{r}{2}-1}^{l-1}\right) + \frac{\Delta r^l g}{4}\left(\rho_r^l + \rho_{r-\myhalf}^l\right).
\end{equation}
We can simplify using
\begin{equation}
\Delta r^{l-1} = 2\Delta r^l,
\end{equation}
\begin{equation}
\rho_{r-\myhalf}^l = \frac{2}{3}\rho_r^l + \frac{1}{3}\rho_{\sfrac{r}{2}-1}^{l-1}.
\end{equation}
Simplifying
\begin{eqnarray}
p_r^l &=& p_{\sfrac{r}{2}-1}^{l-1} + \frac{\Delta r^l g}{2}\left(\frac{2}{3}\rho_r^l + \frac{1}{3}\rho_{\sfrac{r}{2}-1}^{l-1} + \rho_{\sfrac{r}{2}-1}^{l-1}\right) + \frac{\Delta r^l g}{4}\left(\rho_r^l + \frac{2}{3}\rho_r^l + \frac{1}{3}\rho_{\sfrac{r}{2}-1}^{l-1}\right) \nonumber \\
&=& p_{\sfrac{r}{2}-1}^{l-1} + \frac{3\Delta r^l g}{4}\left(\rho_{\sfrac{r}{2}-1}^{l-1} + \rho_r^l\right).\label{Coarse-Fine Stencil}
\end{eqnarray}
When we encounter a change from fine to coarse, analogously
\begin{equation}
p_{(r+1)/2}^{l-1} = p_{r}^l + \frac{3\Delta r^l g}{4}\left(\rho_{r}^l+\rho_{(r+1)/2}^{l-1}\right).\label{Fine-Coarse Stencil}
\end{equation}
\end{document}
